\begin{savequote}[8cm]

	‘When Oldspeak had been once and for all superseded, the last link with the past would have been severed.’
	\qauthor{--- George Orwell, \textit{\usebibentry{orwell1949nineteen}{title}} \citeyearpar{orwell1949nineteen} }

\end{savequote}

\chapter*{\label{ch:preamble}}
\fancyhead[LO]{\emph{Preamble}}
\fancyhead[RE]{\emph{Preamble}}

% Add blank lines before starting
\begin{parse lines}[\noindent]{#1\\}



\end{parse lines}

% In his groundbreaking opus \textit{On the Origin of Species} \citeyearpar{darwin1859origin}, while he was advocating for an evolutionary interpretation of vestigial structures such as the human appendix, Charles Darwin (1809--1882) drew a parallel between the work of linguists and that of evolutionary biologists:
% While he was advocating for an evolutionary interpretation of vestigial structures such as the human appendix in his groundbreaking opus \textit{On the Origin of Species} \citeyearpar{darwin1859origin}, Charles Darwin (1809--1882) drew a parallel between the work of linguists and that of evolutionary biologists:
While Charles Darwin (1809--1882) was advocating an evolutionary interpretation of vestigial structures\footnote{A vestigial structure is an anatomical feature or behaviour that has lost part or all of its initial function and that thus no longer seems to have a purpose in the current species. For instance, the human appendix and coccyx are two such vestigial organs.}
in his groundbreaking opus \textit{On the Origin of Species} \citeyearpar{darwin1859origin}, he drew a parallel between the work of linguists and that of evolutionary biologists:

\begin{quote}
	\textit{‘Rudimentary organs may be compared with the letters in a word, still retained in the spelling, but become useless in the pronunciation, but which serve as a clue in seeking for its derivation.’}
\end{quote}


Nowadays, with the rise of sequencing technologies, the meaningfulness of his analogy is just as topical as ever: evolutionary biologists can now directly ‘read’ DNA and search for its ‘etymology’ by analysing the series of its ‘letters’. 
Ultimately, their goal is to uncover the kinship ties between species, just like linguists would disclose the paths through which words have travelled by examining the remnants of unpronounced letters within them.
% Hopefully, in the end, they could uncover the kinship ties between species, just like linguists could disclose the paths through which words have travelled by examining the remnants of unpronounced letters within them.
% reasons why words have evolved by examining the remnants of letters their relationships.

Indeed, the discovery of DNA in the mid-twentieth century \citep{franklin1953molecular,watson1953molecular,wilkins1953molecular} brought about a real revolution in the study of evolution and even led to the establishment of a new research field to which this thesis belongs: molecular evolution — now rather called evolutionary genomics for whole genomes, rather than single genes, get analysed.
%, for whole genomes can now be analysed.
I will therefore open the introduction in Part~\ref{part:introduction} with Chapter~\ref{ch:1-history-genetics} devoted to tracing back the scientific findings in genetics that directly led to the emergence of this research field aiming at understanding genome evolution.
% this research field aiming at understanding genome evolution
%(Chapter~\ref{ch:1-history-genetics}).
% further progress in the understanding of genome evolution (Chapter~\ref{ch:1-history-genetics}).
% In Chapter~\ref{ch:1-history-genetics}, I will therefore trace back the scientific findings in genetics that directly led to further progress in the understanding of genome evolution.
% in the field of evolutionary biology.

But, precisely, why and how do genomes evolve?
Three main evolutionary forces are classically invoked in this process: mutation, natural selection and genetic drift.
Though, a couple of decades ago, a fourth force made an entrance in the evolutionary scene: biased gene conversion (BGC).
This driver of genome evolution is a direct consequence of recombination — a process essential to meiotic cell division in sexually-reproducing organisms. 
I will thus review the mechanism of meiotic recombination in Chapter~\ref{ch:2-recombination-mechanistics} and the sources of recombination rate variation in Chapter~\ref{ch:3-recombination-variation}. 
This will lead me, in Chapter~\ref{ch:4-gBGC}, to go over the knowledge acquired so far on the fourth evolutionary force of interest for this thesis.

From that point on, I will focus on the puzzling observation which laid the foundation for this work and will set, in Part~\ref{part:objectives}, the objectives we wanted to address.
%goals we wanted to achieve.
% : the fact that the intensity of this force is stable among most species despite the wide range of variation of the parameters on which it depends, and thus

The results presented in Part~\ref{part:results} will then be divided into four chapters.
In Chapter~\ref{ch:5-methodology}, I will describe the unprecedentedly powerful approach we implemented to detect recombination events at high recolution in single individuals.
Next, I will show how we used this method to precisely characterise mouse recombination patterns in Chapter~\ref{ch:6-recombination-parameters} and to quantify biased gene conversion in Chapter~\ref{ch:7-quantification-BGC}.
Last, in Chapter~\ref{ch:8-HFM1}, I will detail how we adapted our method to other studies of recombination with more complex experimental designs involving several genomic introgressions.

Finally, Part~\ref{part:discussion} will be dedicated to discussing this work: I will first consider the scientific implications of our study in Chapter~\ref{ch:9-discussion} and will then share ideas related to it in the broader fields of epistemology, philosophy of science and sociology of knowledge in Chapter~\ref{ch:10-philosophising}.

% Finally, in Part~\ref{part:discussion}, I will discuss the scientific implications of our work (Chapter~\ref{ch:9-discussion}) and will share ideas related to it in the broader fields of epistemology, philosophy of science and sociology of knowledge (Chapter~\ref{ch:10-philosophising}).


% Et, a partir de la, j'expliquerai un des paradoxes (une 'imperfection', un 'vestige' de ce BGC): le fait que l'intensite a l'echelle populationnelle ne varie pas malgre les differences des parametres (taille efficace de la population), ce qui nous amene a poser les objectifs de cette these: mieux comprendre le BGC et ses determinnants au sein des mammiferes (part II)







